Gradual Soundness: Lessons from Static Python

### 2022-05-31 pj-7.1/paper.tex
pj: update ack, fix bib capitalization

%\titlerunning{Draft document---please do not distribute or cite!}
\authorrunning{KC. Lu, B. Greenman, C. Meyer, D. Viehland, A. Panse, and S. Krishnamurthi}

@@

%% 2022-05-27: commented out, from submission
%%\paperdetails{
%%  %% perspective options are: art, sciencetheoretical, scienceempirical, engineering.
%%  perspective=scienceempirical,
%%  %% State one or more areas, separated by a comma. (see 2.2)
%%  %% Please see list of areas in http://programming-journal.org/cfp/
%%  %% The list is open-ended, so use other areas if yours is/are not listed.
%%  area={General-purpose programming},
%%  %% License options include: cc-by (default), cc-by-nc
%%  %license=cc-by,
%%}

@@

\begin{itemize}
  \item
    \emph{Evidence}.
    We present evidence that \SP{} improves performance with few code changes
    / refactorings.
    The Instagram web server has become significantly more efficient
    thanks to a gradual application of \SP{} to high-profile modules.
    Because the server code is closed-source, we additionally present data for
    four public microbenchmarks~(\Cref{s:microbenchmarks}).
    Using only shallow types, \SP{} runs slightly faster than Python
    despite the costs of enforcing soundness~(avg. \pctnum{20}).
    With fine-tuned types, \SP{} consistently outperforms the Python
    baseline~(avg. \pctnum{70}).

@@

To validate the design of \SP{}, we developed a model of the language in PLT
Redex~\cite{kcdeffmrtf-popl-2012}.
The model covers a substantial part of the Python language, including
assertions, loops, exception handlers, and delete statements.
It follows \SP{}'s approach to typing these features.
The model is on GitHub (accessed 2022-05-17):
\begin{center}
  \url{https://github.com/brownplt/insta-model}
\end{center}
Because there is no prior formalization of (all of) Python and of \SP{}, we cannot
\emph{verify} that the model matches them. Instead, we have applied
three methods to give confidence that our model matches reality.
First, we manually reviewed the model in depth---using our
own expertise---to look for non-conformance.  Second, 
we used Redex's property-based testing tools~\cite{kf-sfp-2009} to check
type soundness.
Specifically, we used Redex to generate well-formed terms (using its default
algorithm) and checked whether the well-typed terms that terminated within a
fixed time budget (1,200 expressions and 11,000 programs in a typical run)
produced only well-typed values and allowed errors.
% \footnote{Property-based testing helped us find and fix several issues in the
% model.  None of issues had implications for \SP{}'s soundness.}
Finally, we employed the well-established method of testing end-to-end
conformance with a test suite~\cite{gsk-ecoop-2010,gclpk-dls-2012,pmmwplck-oopsla-2013,bcfgmnss-popl-2014,fgpssmds-popl-2016}.
We translated \numSPtest{} tests from the \SP{} regression suite to the
syntax of the model and confirmed that the results
do match,
which suggests that the model conforms to actual \SP{}.

@@

    \begin{tabular}{l@{\quad}l}
    {Eval. Type $\spteval$} & {Run-time check} \\\hline
    $\sptoptional{\spteval_0}$ & {Accepts either the none value or values that match $\spteval_0$} \\
    $\sptrawpydict$ & {Accepts any Python dictionary} \\
    $\sptchkdict{\spteval_0}{\spteval_1}$ & {Accepts checked dictionaries parameterized by $\spteval_0$ and $\spteval_1$} \\
    $\sptdyn$ & {No check needed, accepts any value} \\
    $\sptclass_0$ & {Accepts instances of the class $\sptclass_0$\tnote{*}} \\
    $\sptnone$ & {Accepts the Python none value} \\
    $\sptint$ & {Accepts integers and booleans} \\
    $\sptbool$ & {Accepts booleans} \\
    $\sptstr$ & {Accepts strings}
    \end{tabular}

@@

We end here, with materialization, rather than present a semantics
for the formalization.
After all, the main benefit of a full formal semantics is to validate
the behavior of complex expressions---and this job is
better left to the Redex mechanization, which supports a much
larger subset of Python~(\Cref{s:surface}).

@@

Interactions with {open-world}~\cite{vss-popl-2017} Python code pose a minor
threat to soundness because
typed functions (and methods) cannot assume that their arguments are
well-typed.
All arguments sent from typed contexts get validated either statically or
via materialization casts, but arguments from untyped contexts are unchecked.
For this reason, \SP{} compiles every typed function to check its inputs.
The optimizer can bypass these checks for typed-to-typed calls~(\Cref{s:optimize}).
Similarly, concrete types such as $\sptrawchkdict$ must check writes that
come from untyped contexts.
%% 2022-01-09 worth mentioning primitive types? they get cast like anything else (they don't materiazize)
%% 2022-01-13 don't think so --- leave prims alone, we don't want to frighten reviews with unsoundness

@@

The Instagram web server has been using \SP{} code to serve requests in
production since April 2021.
Overall, the results are very encouraging.
Instagram's internal profiling tools attribute a \CPUchange{} improvement
in CPU efficiency~(\Cref{s:what-is-cpu}) to \SP{} conversions~(\Cref{s:migration-path}).
This is a big improvement at Instagram scale.

@@

\SP{} first entered the web server codebase as a replacement for a
few critical Cython modules.
Cython had improved performance by compiling these modules to C,
but its partial adoption led to an awkward workflow with a
few compiled modules spread across an interpreted codebase.
Replacing these modules with \SP{} let programmers return to a conventional
Python workflow.
Furthermore, \SP{} types combined with Cinder resulted in a
0.7\% improvement in web server CPU efficiency.
Primitive types~(\Cref{s:c-types}) were essential for matching Cython.

Later \SP{} migrations have been directed by profiling to find frequently-executed code.
During the first half of 2021, the \SP{} team identified hot
modules and proposed types plus small code changes to the maintainers of these modules~(\Cref{s:code-changes}).
The accepted proposals resulted in a 1\% improvement in CPU efficiency.
During the second half of 2021, the \SP{} team applied the same
process at a larger scale.
They also modified a code-generating module to output \SP{} code;
this one change resulted in over four hundred generated typed modules.
Overall, the second-half changes added 2\% to the number of requests
that servers could handle per second at maximum CPU load.

@@

  \centering
  \begin{tabular}{lrr}
    Name               & Typed / Python & Refined / Python \\\midrule
    \bmname{deltablue} &         0.59 &           0.30 \\
    \bmname{fannkuch}  &         1.03 &           0.46 \\
    \bmname{nbody}     &         1.09 &           0.24 \\
    \bmname{richards}  &         0.53 &           0.22
    %% fannkuch typedbasic2 = 0.90
  \end{tabular}

@@

Almost all ratios are less than one, showing that \SP{} achieves a speedup over
Python (typed avg.~$0.8$, refined avg.~$0.3$).
The exceptions are the unrefined versions of \bmname{fannkuch} and
\bmname{nbody}, which is because these benchmarks focus
on number-crunching and \SP{} optimizes arithmetic only for primitive types.
Changing a few types in \bmname{fannkuch} to primitives
gives an immediate speedup (ratio = $0.90$).

@@

\Cref{t:microbenchmark} does not analyze benchmark configurations that mix
typed and untyped code. %% ---only fully-typed and untyped configurations.
A study of mixed configurations (e.g.,~\cite{gm-pepm-2018,gtnffvf-jfp-2019})
may reveal performance bugs that have not appeared in production.

@@

\SP{} is an ambitious undertaking.
Its developers are maintaining a fork of the Python 3.8 runtime,
a type checker,
an optimizing ahead-of-time compiler,
and a method-based JIT compiler.
The team has converted hundreds of untyped modules to \SP{}
and plans to convert thousands more in the future.
And yet, \SP{} is restricted to set attainable goals:
%% for its implementors and for adopters:
\begin{itemize}
  \item
    The syntax of sound static types is relatively small compared
    to \PEP{}.
    Higher-order types are absent.
    The focus is instead on first-order objects and standard data structures~(\Cref{s:ts-context}).
  \item
    The type checker analyzes entire modules rather than individual statements,
    and although it supports a dynamic type, dynamic type-checked code
    does not have the same freedom as untyped code~(\Cref{s:type-dynamic}).
  \item
    Shallow types accommodate Python code but offer coarse
    soundness guarantees.
    %% These types have been the top priority for \SP{}.
    Concrete types are deeply sound, but add nonlocal constraints~(\Cref{s:checked-type}).
    %% \SP{} supports few concrete types at the moment and plans to add more.
  \item
    Gradual class hierarchies are permitted, but restricted to single inheritance.
    Static-to-static method overrides cannot reduce the precision of types~(\Cref{s:inheritance}).
    %% ---a restriction
    %% that enables optimizations but breaks the static gradual guarantee.
  \item
    Progressive primitive types are available to maximize performance~(\Cref{{s:c-types}}).
    %%These types are intentionally incompatible with the dynamic type to help programmers
    %%avoid casts.
\end{itemize}
%
Many of these restrictions go against the grain of the mainstream
research community, but they have been effective in practice.
The current \CPUchange{} increase in Instagram web server CPU efficiency is a huge improvement
and was obtained incrementally without any careful planning to avoid performance bottlenecks.
By contrast, bottlenecks are a major concern for expressive, structurally-typed
gradual languages~\cite{gtnffvf-jfp-2019,tfgnvf-popl-2016}.

@@

\acks{
  \SP{} and the Cinder JIT were developed by
  Max Bernstein,
  John Biesnecker,
  Jacob Bower,
  Sinan Cepel,
  Tiansi Hu,
  Orvid King,
  Vladimir Matveev,
  William Meehan,
  Carl Meyer,
  Matt Page,
  Aniket Panse,
  Brett Simmers,
  Andrei Talaba,
  Dino Viehland,
  and Shiyu Wang.
  The Brown authors contributed to the language by modeling it independently; they
  are not part of the \SP{} team.

  Special thanks to Twitter, through which the two teams
  (specifically, Meyer and Krishnamurthi) met.
  This work was partly supported by Meta and by the US National Science Foundation.
  Greenman is supported by NSF grant 2030859 to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
}

\appendix

@@

We also observe significant untapped opportunity to improve Static Python's
performance; for example, $20\%$ of Typed Fannkuch SP+JIT time is spent on
bounds-checking array accesses; most of this runtime bounds-checking cost could
be eliminated if the Static Python compiler tracks known array sizes and
integer values more thoroughly. And it would also be possible to support list
slicing, list pop, and list insert with primitive arguments, thus avoiding the
need to rewrite these operations in Python for maximum efficiency.

@@

\begin{tabular}{rrrrrr}
  \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} \\
  (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
  $0.57$ & $1.29$ & $7.86$ & $8.36$ & $10.24$ & $16$ \\
  $0.59$ & $1.32$ & $7.92$ & $8.84$ & $9.97$ & $15.73$ \\
  $0.58$ & $1.33$ & $7.89$ & $8.9$ & $10.57$ & $16.2$ \\
  $0.58$ & $1.33$ & $7.92$ & $8.53$ & $10.71$ & $16.71$ \\
  $0.56$ & $1.33$ & $7.83$ & $9.15$ & $9.89$ & $16.62$ \\
  $0.56$ & $1.33$ & $7.97$ & $8.25$ & $10.44$ & $15.65$ \\
  $0.55$ & $1.32$ & $8.18$ & $8.78$ & $10.07$ & $16.22$ \\
  $0.55$ & $1.31$ & $7.85$ & $8.67$ & $10.07$ & $15.73$ \\
  $0.56$ & $1.31$ & $7.96$ & $8.52$ & $10.14$ & $15.82$ \\
  $0.57$ & $1.32$ & $8$ & $8.35$ & $10.57$ & $16.66$ \\
\end{tabular}

\begin{tabular}{rrrrrr}
  \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} \\
  (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
  $1.41$ & $2.68$ & $7.67$ & $5.58$ & $7.97$ & $13.32$ \\
  $1.38$ & $2.6$ & $7.9$ & $5.92$ & $7.25$ & $13.59$ \\
  $1.39$ & $2.64$ & $7.74$ & $5.54$ & $7.64$ & $13.83$ \\
  $1.38$ & $2.65$ & $7.8$ & $6.06$ & $7.24$ & $13.95$ \\
  $1.37$ & $2.62$ & $7.68$ & $5.52$ & $7.2$ & $13.5$ \\
  $1.39$ & $2.58$ & $7.8$ & $5.63$ & $7.27$ & $13.52$ \\
  $1.36$ & $2.65$ & $7.79$ & $5.74$ & $7.17$ & $13.34$ \\
  $1.43$ & $2.66$ & $7.82$ & $5.56$ & $7.54$ & $13.68$ \\
  $1.53$ & $2.63$ & $7.9$ & $5.59$ & $6.89$ & $13.79$ \\
  $1.37$ & $2.71$ & $7.95$ & $5.74$ & $6.91$ & $13.5$ \\
\end{tabular}

\begin{tabular}{rrr}
  \colname{Orig} & \colname{Orig} & \colname{Orig} \\
  (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
  $2.6$ & $3.88$ & $10.17$ \\
  $2.6$ & $4.01$ & $10.18$ \\
  $2.61$ & $4.17$ & $10.73$ \\
  $2.9$ & $3.92$ & $9.96$ \\
  $2.63$ & $3.79$ & $10.04$ \\
  $2.64$ & $3.95$ & $9.99$ \\
  $2.62$ & $3.9$ & $10.16$ \\
  $2.6$ & $3.84$ & $9.69$ \\
  $2.6$ & $4.04$ & $9.82$ \\
  $2.62$ & $3.85$ & $9.85$ \\
\end{tabular}

@@

  \begin{tabular}{rrrrrr}
  \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} \\
  (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
  $1.27$ & $1.39$ & $34.69$ & $40.74$ & $41$ & $48.23$ \\
  $1.32$ & $1.25$ & $34.35$ & $40.23$ & $40.94$ & $49.03$ \\
  $1.28$ & $1.26$ & $34.67$ & $40.74$ & $40.65$ & $49.42$ \\
  $1.23$ & $1.31$ & $35.71$ & $40.94$ & $39.7$ & $49.35$ \\
  $1.28$ & $1.38$ & $34.85$ & $39.59$ & $40.46$ & $49.18$ \\
  $1.26$ & $1.34$ & $33.96$ & $40.45$ & $39.99$ & $47.79$ \\
  $1.26$ & $1.23$ & $34.65$ & $40.78$ & $40.97$ & $49.7$ \\
  $1.27$ & $1.35$ & $33.83$ & $40.96$ & $41.78$ & $47.92$ \\
  $1.23$ & $1.4$ & $36.23$ & $40.49$ & $40.29$ & $48.68$ \\
  $1.23$ & $1.27$ & $36.05$ & $40.72$ & $40.11$ & $48.29$ \\
\end{tabular}

\begin{tabular}{rrrrrr}
  \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} \\
  (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
  $2.77$ & $2.83$ & $6.18$ & $2.89$ & $2.86$ & $4.07$ \\
  $2.85$ & $2.8$ & $6.12$ & $2.94$ & $2.9$ & $4.06$ \\
  $2.79$ & $2.77$ & $6.38$ & $2.87$ & $2.86$ & $4.1$ \\
  $2.87$ & $2.81$ & $6.23$ & $2.87$ & $2.85$ & $4.29$ \\
  $2.85$ & $2.84$ & $6.17$ & $2.92$ & $2.81$ & $4.14$ \\
  $2.87$ & $2.8$ & $6.21$ & $2.94$ & $2.8$ & $4.06$ \\
  $2.92$ & $2.76$ & $6.04$ & $3.11$ & $3.05$ & $4.08$ \\
  $2.78$ & $2.76$ & $6.13$ & $3.1$ & $2.86$ & $4.19$ \\
  $2.84$ & $2.76$ & $6.13$ & $2.87$ & $2.96$ & $4.31$ \\
  $2.92$ & $2.79$ & $6.17$ & $2.91$ & $2.96$ & $4.1$ \\
\end{tabular}

>>>>

%\titlerunning{Draft document---please do not distribute or cite!}
\authorrunning{K.-C. Lu, B. Greenman, C. Meyer, D. Viehland, A. Panse, S. Krishnamurthi}

@@

%%\paperdetails{
%%  %% perspective options are: art, sciencetheoretical, scienceempirical, engineering.
%%  perspective=scienceempirical,
%%  %% State one or more areas, separated by a comma. (see 2.2)
%%  %% Please see list of areas in http://programming-journal.org/cfp/
%%  %% The list is open-ended, so use other areas if yours is/are not listed.
%%  area={General-purpose programming},
%%  %% License options include: cc-by (default), cc-by-nc
%%  %license=cc-by,
%%}

@@

\begin{itemize}
  \item
    \emph{Evidence}.
    We present evidence that \SP{} improves performance with few code changes
    / refactorings.
    The Instagram web server has become significantly more efficient
    thanks to a gradual application of \SP{} to high-profile modules.
    Because the server code is closed-source, we additionally present data for
    four public microbenchmarks~(\Cref{s:microbenchmarks}).
    Using only shallow types, \SP{} runs slightly faster than Python
    despite the costs of enforcing soundness~(avg.~\pctnum{20}).
    With fine-tuned types, \SP{} consistently outperforms the Python
    baseline~(avg.~\pctnum{70}).

@@

To validate the design of \SP{}, we developed a model of the language in PLT
Redex~\cite{kcdeffmrtf-popl-2012}.
The model covers a substantial part of the Python language, including
assertions, loops, exception handlers, and delete statements.
It follows \SP{}'s approach to typing these features.
The model is on GitHub (accessed 2022-05-17):
\begin{center}
  \url{https://github.com/brownplt/insta-model}
\end{center}
Because there is no prior formalization of (all of) Python and of \SP{}, we cannot
\emph{verify} that the model matches them. Instead, we have applied
three methods to give confidence that our model matches reality.
First, we manually reviewed the model in depth\--- using our
own expertise\--- to look for non-conformance.  Second, 
we used Redex's property-based testing tools~\cite{kf-sfp-2009} to check
type soundness.
Specifically, we used Redex to generate well-formed terms (using its default
algorithm) and checked whether the well-typed terms that terminated within a
fixed time budget (1,200 expressions and 11,000 programs in a typical run)
produced only well-typed values and allowed errors.
% \footnote{Property-based testing helped us find and fix several issues in the
% model.  None of issues had implications for \SP{}'s soundness.}
Finally, we employed the well-established method of testing end-to-end
conformance with a test suite~\cite{gsk-ecoop-2010,gclpk-dls-2012,pmmwplck-oopsla-2013,bcfgmnss-popl-2014,fgpssmds-popl-2016}.
We translated \numSPtest{} tests from the \SP{} regression suite to the
syntax of the model and confirmed that the results
do match,
which suggests that the model conforms to actual \SP{}.

@@

    \begin{tabular}{l@{\quad}l}
    \toprule
    {Eval. Type $\spteval$} & {Run-time check} \\\midrule
    $\sptoptional{\spteval_0}$ & {Accepts either the none value or values that match $\spteval_0$} \\
    $\sptrawpydict$ & {Accepts any Python dictionary} \\
    $\sptchkdict{\spteval_0}{\spteval_1}$ & {Accepts checked dictionaries parameterized by $\spteval_0$ and $\spteval_1$} \\
    $\sptdyn$ & {No check needed, accepts any value} \\
    $\sptclass_0$ & {Accepts instances of the class $\sptclass_0$\tnote{*}} \\
    $\sptnone$ & {Accepts the Python none value} \\
    $\sptint$ & {Accepts integers and booleans} \\
    $\sptbool$ & {Accepts booleans} \\
    $\sptstr$ & {Accepts strings} \\
    \bottomrule
    \end{tabular}

@@

We end here, with materialization, rather than present a semantics
for the formalization.
After all, the main benefit of a full formal semantics is to validate
the behavior of complex expressions\--- and this job is
better left to the Redex mechanization, which supports a much
larger subset of Python~(\Cref{s:surface}).

@@

Interactions with {open-world}~\cite{vss-popl-2017} Python code pose a minor
threat to soundness because
typed functions (and methods) cannot assume that their arguments are
well-typed.
All arguments sent from typed contexts get validated either statically or
via materialization casts, but arguments from untyped contexts are unchecked.
For this reason, \SP{} compiles every typed function to check its inputs.
The optimizer can bypass these checks for typed-to-typed calls~(\Cref{s:optimize}).
Similarly, concrete types such as $\sptrawchkdict$ must check writes that
come from untyped contexts.
%% 2022-01-09 worth mentioning primitive types? they get cast like anything else (they don't materiazize)
%% 2022-01-13 don't think so \---  leave prims alone, we don't want to frighten reviews with unsoundness

@@

The Instagram web server has been using \SP{} code to serve requests in
production since April 2021.
Overall, the results are very encouraging.
Instagram's internal profiling tools attribute a \CPUchange{} improvement
in CPU efficiency~(\Cref{s:what-is-cpu})~to \SP{} conversions~(\Cref{s:migration-path}).
This is a big improvement at Instagram scale.

@@

\SP{} first entered the web server codebase as a replacement for a
few critical Cython modules.
Cython had improved performance by compiling these modules to C,
but its partial adoption led to an awkward workflow with a
few compiled modules spread across an interpreted codebase.
Replacing these modules with \SP{} let programmers return to a conventional
Python workflow.
Furthermore, \SP{} types combined with Cinder resulted in a
\SI{0.7}{\percent} improvement in web server CPU efficiency.
Primitive types~(\Cref{s:c-types}) were essential for matching Cython.

Later \SP{} migrations have been directed by profiling to find frequently-executed code.
During the first half of 2021, the \SP{} team identified hot
modules and proposed types plus small code changes to the maintainers of these modules~(\Cref{s:code-changes}).
The accepted proposals resulted in a \SI{1}{\percent} improvement in CPU efficiency.
During the second half of 2021, the \SP{} team applied the same
process at a larger scale.
They also modified a code-generating module to output \SP{} code;
this one change resulted in over four hundred generated typed modules.
Overall, the second-half changes added \SI{2}{\percent} to the number of requests
that servers could handle per second at maximum CPU load.

@@

  \centering
  \begin{tabular}{lSS}\toprule
    Name               & {Typed / Python} & {Refined / Python} \\\midrule
    \bmname{deltablue} &         0.59 &           0.30 \\
    \bmname{fannkuch}  &         1.03 &           0.46 \\
    \bmname{nbody}     &         1.09 &           0.24 \\
    \bmname{richards}  &         0.53 &           0.22 \\
    \bottomrule    
    %% fannkuch typedbasic2 = 0.90
  \end{tabular}

@@

Almost all ratios are less than one, showing that \SP{} achieves a speedup over
Python (typed avg.~\num{0.8}, refined avg.~\num{0.3}).
The exceptions are the unrefined versions of \bmname{fannkuch} and
\bmname{nbody}, which is because these benchmarks focus
on number-crunching and \SP{} optimizes arithmetic only for primitive types.
Changing a few types in \bmname{fannkuch} to primitives
gives an immediate speedup (ratio = \num{0.90}).

@@

\Cref{t:microbenchmark} does not analyze benchmark configurations that mix
typed and untyped code. %% \--- only fully-typed and untyped configurations.
A study of mixed configurations (e.g.,~\cite{gm-pepm-2018,gtnffvf-jfp-2019})
may reveal performance bugs that have not appeared in production.

@@

\SP{} is an ambitious undertaking.
Its developers are maintaining a fork of the Python 3.8 runtime,
a type checker,
an optimizing ahead-of-time compiler,
and a method-based JIT compiler.
The team has converted hundreds of untyped modules to \SP{}
and plans to convert thousands more in the future.
And yet, \SP{} is restricted to set attainable goals:
%% for its implementors and for adopters:
\begin{itemize}
  \item
    The syntax of sound static types is relatively small compared
    to \PEP{}.
    Higher-order types are absent.
    The focus is instead on first-order objects and standard data structures~(\Cref{s:ts-context}).
  \item
    The type checker analyzes entire modules rather than individual statements,
    and although it supports a dynamic type, dynamic type-checked code
    does not have the same freedom as untyped code~(\Cref{s:type-dynamic}).
  \item
    Shallow types accommodate Python code but offer coarse
    soundness guarantees.
    %% These types have been the top priority for \SP{}.
    Concrete types are deeply sound, but add nonlocal constraints~(\Cref{s:checked-type}).
    %% \SP{} supports few concrete types at the moment and plans to add more.
  \item
    Gradual class hierarchies are permitted, but restricted to single inheritance.
    Static-to-static method overrides cannot reduce the precision of types~(\Cref{s:inheritance}).
    %% \--- a restriction
    %% that enables optimizations but breaks the static gradual guarantee.
  \item
    Progressive primitive types are available to maximize performance~(\Cref{{s:c-types}}).
    %%These types are intentionally incompatible with the dynamic type to help programmers
    %%avoid casts.
\end{itemize}
%
Many of these restrictions go against the grain of the mainstream
research community, but they have been effective in practice.
The current \CPUchange{} increase in Instagram web server CPU efficiency is a huge improvement
and was obtained incrementally without any careful planning to avoid performance bottlenecks.
By contrast, bottlenecks are a major concern for expressive, structurally-typed
gradual languages~\cite{gtnffvf-jfp-2019,tfgnvf-popl-2016}.

@@

\acks
  \SP{} and the Cinder JIT were developed by
  Max Bernstein,
  John Biesnecker,
  Jacob Bower,
  Sinan Cepel,
  Tiansi Hu,
  Orvid King,
  Vladimir Matveev,
  William Meehan,
  Carl Meyer,
  Matt Page,
  Aniket Panse,
  Brett Simmers,
  Andrei Talaba,
  Dino Viehland,
  and Shiyu Wang.
  The Brown authors contributed to the language by modeling it independently; they
  are not part of the \SP{} team.

  Special thanks to Twitter, through which the two teams
  (specifically, Meyer and Krishnamurthi) met.
  Thanks also to Tobias Pape for going above and beyond to offer splendid assistance with the paper typesetting.
  This work was partly supported by Meta and by the US National Science Foundation.
  Greenman is supported by NSF grant 2030859 to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.

\clearpage
\appendix

@@

We also observe significant untapped opportunity to improve Static Python's
performance; for example, \SI{20}{\percent} of Typed Fannkuch SP+JIT time is spent on
bounds-checking array accesses; most of this runtime bounds-checking cost could
be eliminated if the Static Python compiler tracks known array sizes and
integer values more thoroughly. And it would also be possible to support list
slicing, list pop, and list insert with primitive arguments, thus avoiding the
need to rewrite these operations in Python for maximum efficiency.

@@

\begin{tabular}{SSSSSS}
  \toprule
  \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} \\
  {(\colname{SP} \colname{JIT} \colname{SF})} & {(\colname{SP} \colname{JIT})} & {(\colname{SP})} & {(\colname{JIT} \colname{SF})} & {(\colname{JIT})} & {()} \\\midrule
  0.57 & 1.29 & 7.86 & 8.36 & 10.24 & 16 \\
  0.59 & 1.32 & 7.92 & 8.84 & 9.97 & 15.73 \\
  0.58 & 1.33 & 7.89 & 8.9 & 10.57 & 16.2 \\
  0.58 & 1.33 & 7.92 & 8.53 & 10.71 & 16.71 \\
  0.56 & 1.33 & 7.83 & 9.15 & 9.89 & 16.62 \\
  0.56 & 1.33 & 7.97 & 8.25 & 10.44 & 15.65 \\
  0.55 & 1.32 & 8.18 & 8.78 & 10.07 & 16.22 \\
  0.55 & 1.31 & 7.85 & 8.67 & 10.07 & 15.73 \\
  0.56 & 1.31 & 7.96 & 8.52 & 10.14 & 15.82 \\\
  0.57 & 1.32 & 8 & 8.35 & 10.57 & 16.66 \\
  \bottomrule
\end{tabular}

\begin{tabular}{SSSSSS}
  \toprule
  \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} \\
  {(\colname{SP} \colname{JIT} \colname{SF})} & {(\colname{SP} \colname{JIT})} & {(\colname{SP})} & {(\colname{JIT} \colname{SF})} & {(\colname{JIT})} & {()} \\\midrule
  1.41 & 2.68 & 7.67 & 5.58 & 7.97 & 13.32 \\
  1.38 & 2.6 & 7.9 & 5.92 & 7.25 & 13.59 \\
  1.39 & 2.64 & 7.74 & 5.54 & 7.64 & 13.83 \\
  1.38 & 2.65 & 7.8 & 6.06 & 7.24 & 13.95 \\
  1.37 & 2.62 & 7.68 & 5.52 & 7.2 & 13.5 \\
  1.39 & 2.58 & 7.8 & 5.63 & 7.27 & 13.52 \\
  1.36 & 2.65 & 7.79 & 5.74 & 7.17 & 13.34 \\
  1.43 & 2.66 & 7.82 & 5.56 & 7.54 & 13.68 \\
  1.53 & 2.63 & 7.9 & 5.59 & 6.89 & 13.79 \\
  1.37 & 2.71 & 7.95 & 5.74 & 6.91 & 13.5 \\
  \bottomrule
\end{tabular}

\begin{tabular}{SSS}
  \toprule
  \colname{Orig} & \colname{Orig} & \colname{Orig} \\
  {(\colname{JIT} \colname{SF})} & {(\colname{JIT})} & {()} \\\midrule
  2.6 & 3.88 & 10.17 \\
  2.6 & 4.01 & 10.18 \\
  2.61 & 4.17 & 10.73 \\
  2.9 & 3.92 & 9.96 \\
  2.63 & 3.79 & 10.04 \\
  2.64 & 3.95 & 9.99 \\
  2.62 & 3.9 & 10.16 \\
  2.6 & 3.84 & 9.69 \\
  2.6 & 4.04 & 9.82 \\
  2.62 & 3.85 & 9.85 \\
  \bottomrule
\end{tabular}

@@

  \begin{tabular}{SSSSSS}
  \toprule
  \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} \\
  {(\colname{SP} \colname{JIT} \colname{SF})} & {(\colname{SP} \colname{JIT})} & {(\colname{SP})} & {(\colname{JIT} \colname{SF})} & {(\colname{JIT})} & {()} \\\midrule
  1.27 & 1.39 & 34.69 & 40.74 & 41 & 48.23 \\
  1.32 & 1.25 & 34.35 & 40.23 & 40.94 & 49.03 \\
  1.28 & 1.26 & 34.67 & 40.74 & 40.65 & 49.42 \\
  1.23 & 1.31 & 35.71 & 40.94 & 39.7 & 49.35 \\
  1.28 & 1.38 & 34.85 & 39.59 & 40.46 & 49.18 \\
  1.26 & 1.34 & 33.96 & 40.45 & 39.99 & 47.79 \\
  1.26 & 1.23 & 34.65 & 40.78 & 40.97 & 49.7 \\
  1.27 & 1.35 & 33.83 & 40.96 & 41.78 & 47.92 \\
  1.23 & 1.4 & 36.23 & 40.49 & 40.29 & 48.68 \\
  1.23 & 1.27 & 36.05 & 40.72 & 40.11 & 48.29 \\
  \bottomrule
\end{tabular}

\begin{tabular}{SSSSSS}
  \toprule
  \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} \\
  {(\colname{SP} \colname{JIT} \colname{SF})} & {(\colname{SP} \colname{JIT})} & {(\colname{SP})} & {(\colname{JIT} \colname{SF})} & {(\colname{JIT})} & {()} \\\midrule
  2.77 & 2.83 & 6.18 & 2.89 & 2.86 & 4.07 \\
  2.85 & 2.8 & 6.12 & 2.94 & 2.9 & 4.06 \\
  2.79 & 2.77 & 6.38 & 2.87 & 2.86 & 4.1 \\
  2.87 & 2.81 & 6.23 & 2.87 & 2.85 & 4.29 \\
  2.85 & 2.84 & 6.17 & 2.92 & 2.81 & 4.14 \\
  2.87 & 2.8 & 6.21 & 2.94 & 2.8 & 4.06 \\
  2.92 & 2.76 & 6.04 & 3.11 & 3.05 & 4.08 \\
  2.78 & 2.76 & 6.13 & 3.1 & 2.86 & 4.19 \\
  2.84 & 2.76 & 6.13 & 2.87 & 2.96 & 4.31 \\
  2.92 & 2.79 & 6.17 & 2.91 & 2.96 & 4.1 \\
  \bottomrule
\end{tabular}

\begin{tabular}{SSSSSS}
  \toprule
  \colname{T-Min-2} & \colname{T-Min-2} & \colname{T-Min-2} & \colname{T-Min-2} & \colname{T-Min-2} & \colname{T-Min-2} \\
  {(\colname{SP} \colname{JIT} \colname{SF})} & {(\colname{SP} \colname{JIT})} & {(\colname{SP})} & {(\colname{JIT} \colname{SF})} & {(\colname{JIT})} & {()} \\\midrule
  2.48 & 2.52 & 7.28 & 3.16 & 3.08 & 4.59 \\
  2.52 & 2.48 & 7.28 & 3.26 & 3.08 & 4.51 \\
  2.47 & 2.53 & 7.24 & 3.11 & 3.18 & 4.59 \\
  2.48 & 2.57 & 7.35 & 3.18 & 3.14 & 4.73 \\
  2.5 & 2.56 & 7.34 & 3.14 & 3.12 & 4.63 \\
  2.43 & 2.47 & 7.17 & 3.23 & 3.09 & 4.53 \\
  2.47 & 2.69 & 7.18 & 3.41 & 3.21 & 4.48 \\
  2.47 & 2.45 & 7.14 & 3.43 & 3.14 & 4.53 \\
  2.43 & 2.55 & 7.23 & 3.08 & 3.24 & 4.53 \\
  2.52 & 2.49 & 7.2 & 3.16 & 3.18 & 4.45 \\
  \bottomrule
\end{tabular}
